Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-07-10 final/checking-executions.tex
half of the paper

%TODO REPHRASE THIS
%While checking linearizability of a given execution is known to be NP-complete in general~\cite{journals/siamcomp/GibbonsK97} we show in this section that it is polynomial-time for data-differentiated executions of the priority queue.
We define a recursive procedure for checking linearizability of an execution w.r.t. $\seqPQ$.
To ease the exposition, Section~\ref{ssec:seq_exec} introduces a recursive procedure for checking whether a data-differentiated \emph{sequential} execution is admitted by the priority queue which is then extended to the concurrent case in Section~\ref{ssec:conc_exec}.

@@

Checking whether a data-differentiated sequential execution belongs to $\seqPQ$ could be implemented by checking membership into {\color {blue}the language accepted by the LTS $PQ$}{\color {red}the set of traces of the LTS $PQ$}. The recursive procedure $\textit{Check-PQ-Seq}$ outlined in Algorithm~\ref{alg:seq_check} is an alternative to this membership test. Roughly, it selects one or two operations in the input execution, checks whether their return values are correct by ignoring the order between the other operations other than how they are ordered w.r.t. the selected ones, and calls itself recursively on the execution without the selected operations. %{\color {blue} Here we implicitly mix operation $o$ (when $\textit{call}_o(m,a)$, $\textit{ret}_o(m,b)$ in $e$) with $m(a,b)$.}

We explain how the procedure works on the following execution:
{\color {red}
\begin{align}
\hspace{-5mm}\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{rm}(\textit{empty}) \cdot \textit{put}(d,p_2) \cdot \textit{put}(f,p_3) \cdot \textit{rm}(f) \cdot \textit{put}(b,p_1)\label{eq:ex_rec1}
\end{align}
}
where $p_1$, $p_2$, $p_3$ are priorities such that $p_1 \prec p_2$ and $p_1 \prec p_3$, and $p_2$ and $p_3$ are incomparable. Since the $\textit{rm}(\textit{empty})$ operations %\footnote{An $m(a)$-operation in an execution $e$ is an operation identifier $o$ s.t. $e$ contains the actions $\textit{call}_o(m,a)$, $\textit{ret}_o(m,a)$.}
are read-only (they don't affect the state of the queue), they are selected first. Ensuring that an operation $o=\textit{rm}(\textit{empty})$ is correct boils down to checking that every $\textit{put}(x,p)$ operation before $o$ is matched to a $\textit{rm}(x)$ operation which also occurs before $o$. This is true in this case for $x\in \{a,c\}$. Therefore, the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of
{\color {red}
\begin{align*}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(f,p_3) \cdot \textit{rm}(f) \cdot \textit{put}(b,p_1)
\end{align*}
}
When there are no more $\textit{rm}(\textit{empty})$ operations, the procedure selects a $\textit{put}$ operation adding a value with maximal priority which is not removed, and then a pair of $\textit{put}$ and $\textit{rm}$ operations adding and removing the same maximal priority value. For instance, since $p_2$ is a maximal priority, it selects the operation $\textit{put}(d,p_2)$. {This operation is correct if $d$ is the last value with priority $p_2$, and the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of
{\color {red}
\begin{align*}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(f,p_3) \cdot \textit{rm}(f) \cdot \textit{put}(b,p_1)
\end{align*}
}
Since there is no other value of maximal priority which is not removed, the procedure selects a pair of $\textit{put}$/$\textit{rm}$ operations with an argument of maximal priority $p_2$,} for instance, $\textit{put}(c,p_2)$ and $\textit{rm}(c)$. The value returned by $\textit{rm}$ is correct if all the values of priority smaller than $p_2$ added before $\textit{rm}(c)$ are also removed before $\textit{rm}(c)$. In this case, $a$ is the only value of priority smaller than $p_2$ and it satisfies this property. Applying a similar reasoning for all the remaining values, it can be proved that this execution is correct.

\begin{algorithm}[t]
\KwIn {A data-differentiated sequential execution $e$}
\KwOut{$\mathsf{true}$ iff $e\in \seqPQ$}

@@

\Else
{
    \If {$\exists\ x \in \textit{values}(e)$ such that $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ holds}
    {
        \KwRet $\textit{Check-PQ-Seq}(e \setminus x)$\;
    }
    \Else {\KwRet $\mathsf{false}$\;}
}
\caption{$\textit{Check-PQ-Seq}$}
\label{alg:seq_check}
\end{algorithm}

In formal terms, the operations which are selected depend on the following set of predicates on executions:
{\small
\begin{align*}
\mathsf{Has\text{-}EmptyRemoves}(e)=\mathsf{true} & \mbox{ iff  $e$ contains a $\textit{rm}(\textit{empty})$ operation} \\
\mathsf{Has\text{-}UnmatchedMaxPriority}(e)=\mathsf{true} & \mbox{ iff $p\in \textit{unmatched-priorities}(e)$ for a maximal priority} \\
&\hspace{4mm}\mbox{$p\in priorities(e)$}
\end{align*}}
{\color{blue} where $\textit{priorities}(l)$, resp., $\textit{unmatched-priorities}(l)$}{\color{red} where $\textit{priorities}(e)$, resp., $\textit{unmatched-priorities}(e)$}, is the set of priorities occurring in $\textit{put}$ operations of $e$, resp., in $\textit{put}$ operations of $e$ for which there is no $\textit{rm}$ operation removing the same value. We call the latter \emph{unmatched} put operations. A put operation which is not unmatched is called \emph{matched}. For simplicity, we consider the following syntactic sugar $\mathsf{Has\text{-}MatchedMaxPriority}(e)=\neg \mathsf{Has\text{-}EmptyRemoves}(e)\land \neg \mathsf{Has\text{-}UnmatchedMaxPriority}(e)$. By an abuse of notation, we also assume that  $\mathsf{Has\text{-}UnmatchedMaxPriority}(e) \Rightarrow \neg \mathsf{Has\text{-}EmptyRemoves}(e)$ (this is sound by the order of the conditionals in $\textit{Check-PQ-Seq}$).

The predicates defining the correctness of the selected operations are defined as follows:
{\small
\begin{align*}
\mathsf{EmptyRemove\text{-}Seq}(e,o)=\mathsf{true} & \mbox{ iff  $e= u\cdot o\cdot v$ and $\textit{matched}(u)$} \\
\mathsf{UnmatchedMaxPriority\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $e= u\cdot \textit{put}(x,p)\cdot v$, $p\not\prec \textit{priorities}(u\cdot v)$, and $p\not\in \textit{priorities}(v)$} \\
\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $e= u\cdot \textit{put}(x,p)\cdot v\cdot \textit{rm}(x)\cdot w$, $p\not\prec \textit{priorities}(u\cdot v\cdot w)$,} \\
&\hspace{4mm}\mbox{$p\not\preceq \textit{unmatched-priorities}(u\cdot v\cdot w)$, $\textit{matched}_\prec(u\cdot v,p)$,} \\
&\hspace{4mm}\mbox{and $p\not\in \textit{priorities}(v\cdot w)$}
\end{align*}
}

\vspace{-3mm}
\noindent
where $p\prec \textit{priorities}(e)$ when $p\prec p'$ for some $p'\in \textit{priorities}(e)$ (and similarly for $p\prec \textit{unmatched-priorities}(e)$ or $p\preceq \textit{unmatched-priorities}(e)$),
$\textit{matched}_\prec(e,p)$ holds when each value with priority strictly smaller than $p$ is removed in $e$, and $\textit{matched}(e)$ holds when $\textit{matched}_\prec(e,p)$ holds for each $p\in\mathbb{P}$. Compared to the example presented at the beginning of the section, these predicates take into consideration that multiple values with the same priority are removed in FIFO order: the predicate $\mathsf{MatchedMaxPrioritySeq}(e,x)$ holds when $x$ is the last value with priority $p$ added in $e$.

When $o$ is a $\textit{rm}(\textit{empty})$ operation, we write $e\setminus o$ for the maximal subsequence of $e$ which doesn't contain $o$. For an execution $e$, $\textit{values}(e)$ is the set of values occurring in call/return actions of $e$.

@@

The extension of $\textit{Check-PQ-Seq}$ to concurrent executions, checking whether they are linearizable w.r.t. $\seqPQ$, is obtained by replacing every predicate $\Gamma\mathsf{\text{-}Seq}$ with
\begin{align*}
\Gamma\mathsf{\text{-}Conc}(e,\alpha) = \mathsf{true}\mbox{ iff there exists a sequential execution $s$ such that $e\sqsubseteq s$ and $\Gamma\mathsf{\text{-}Seq}(s,\alpha)$}
\end{align*}
for each $\Gamma\in \{\mathsf{EmptyRemove}, \mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$. Let $\textit{Check-PQ-Conc}$ denote the thus obtained procedure (we assume recursive calls are modified accordingly).

@@

The following lemma states the correctness of $\textit{Check-PQ-Conc}$. Completeness follows easily from the properties of $\seqPQ$. Thus, if $\textit{Check-PQ-Conc}(e) = \textit{false}$, then there exists a set $D$ of values s.t. either  $\mathsf{EmptyRemove\text{-}Conc}(e \vert D)$ is false, or $\mathsf{UnmatchedMaxPriority\text{-}Conc}(e \vert D,x)$ is false for all the values $x$ of maximal priority that are not removed (and there exists at least one such value), or $\mathsf{MatchedMaxPriority\text{-}Conc}(e \vert D,x)$ is false for all the values $x$ of maximal priority (and these values are all removed in $e \vert D$). It can be easily seen that we get $e \vert D\not\sqsubseteq \seqPQ$ in all cases, which by the closure under projection of $\seqPQ$ implies, $e \not\sqsubseteq \seqPQ$ (since every linearization of $e$ includes as a subsequence a linearization of $e \vert D$).

\begin{restatable}{lemma}{ConCheckEPQIsCorrect}
\label{lemma:con-check-EPQ is correct}
$\textit{Check-PQ-Conc}(e)=\mathsf{true}$ iff $e \sqsubseteq \seqPQ$, for every data-differentiated execution $e$.
\end{restatable}

@@

%This completes the proof of this lemma.
%
%
%\qed
%\end {proof}
\begin{figure}[t]
  \centering
  \includegraphics[width=1 \textwidth]{figures/PIC-HIS-EPQ1-TwoHis-2.pdf}
%\vspace{-10pt}
  \caption{The process of obtaining linearization of $e$}
  \label{fig:concurrent execution for EPQ1}
\end{figure}

>>>>

%TODO REPHRASE THIS
%While checking linearizability of a given execution is known to be NP-complete in general~\cite{journals/siamcomp/GibbonsK97} we show in this section that it is polynomial-time for data-differentiated executions of the priority queue.
We define a recursive procedure for checking linearizability of a data-differentiated execution w.r.t. $\seqPQ$.
To ease the exposition, Section~\ref{ssec:seq_exec} introduces a recursive procedure for checking whether a data-differentiated \emph{sequential} execution is admitted by the priority queue which is then extended to the concurrent case in Section~\ref{ssec:conc_exec}.

@@

%Checking whether a data-differentiated sequential execution belongs to $\seqPQ$ could be implemented by checking membership into {\color {blue}the language accepted by the LTS $PQ$}{\color {red}the set of traces of the LTS $PQ$}. 
The recursive procedure $\textit{Check-PQ-Seq}$ outlined in Algorithm~\ref{alg:seq_check} checks whether a data-differentiated sequential execution belongs to $\seqPQ$ (i.e., if it is accepted by the LTS $PQ$).
%is an alternative to this membership test. 
 Roughly, it selects one or two operations in the input execution, checks whether their return values are correct by ignoring the order between the other operations other than how they are ordered w.r.t. the selected ones, and calls itself recursively on the execution without the selected operations. %{\color {blue} Here we implicitly mix operation $o$ (when $\textit{call}_o(m,a)$, $\textit{ret}_o(m,b)$ in $e$) with $m(a,b)$.}

We explain how the procedure works on the following execution:
\begin{align}
\hspace{-5mm}\textit{put}(c,p_2)\hspace{-.5mm} \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{rm}(\textit{empty}) \cdot \textit{put}(d,p_2) \cdot \textit{put}(\mathit{f},p_3) \cdot \textit{rm}(\mathit{f}) \cdot \textit{put}(b,p_1)\label{eq:ex_rec1}
\end{align}
where $p_1$, $p_2$, $p_3$ are priorities such that $p_1 \prec p_2$ and $p_1 \prec p_3$, and $p_2$ and $p_3$ are incomparable. Since the $\textit{rm}(\textit{empty})$ operations %\footnote{An $m(a)$-operation in an execution $e$ is an operation identifier $o$ s.t. $e$ contains the actions $\textit{call}_o(m,a)$, $\textit{ret}_o(m,a)$.}
are read-only (they don't affect the queue's state), they are selected first. An $\textit{rm}(\textit{empty})$-operation $o$ is correct when every $\textit{put}(x,p)$ operation before $o$ is matched to a $\textit{rm}(x)$ operation which also occurs before $o$. This is true in this case for $x\in \{a,c\}$. Thus, the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of
\begin{align}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(f,p_3) \cdot \textit{rm}(f) \cdot \textit{put}(b,p_1)\label{eq:ex_rec2}
\end{align}
When the execution contains no $\textit{rm}(\textit{empty})$-operation, the procedure selects a $\textit{put}$ operation adding a value that is not removed and that has a maximal priority. For (\ref{eq:ex_rec2}), it selects $\textit{put}(d,p_2)$ because $p_2$ is a maximal priority. This operation is correct since $d$ is the last value with priority $p_2$ in the execution, and the correctness of (\ref{eq:ex_rec2}) reduces to the correctness of
\begin{align}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(f,p_3) \cdot \textit{rm}(f) \cdot \textit{put}(b,p_1)\label{eq:ex_rec3}
\end{align}
If no operations like above can be found, $\textit{Check-PQ-Seq}$ selects a pair of $\textit{put}$ and $\textit{rm}$ operations adding and removing the same maximal priority value. For (\ref{eq:ex_rec2}), it can select 
%Since there is no other value of maximal priority which is not removed, the procedure selects a pair of $\textit{put}$/$\textit{rm}$ operations with an argument of maximal priority $p_2$, for instance, 
$\textit{put}(c,p_2)$ and $\textit{rm}(c)$. The value returned by $\textit{rm}(c)$ is correct if all the values of priority smaller than $p_2$ added before $\textit{rm}(c)$ are also removed before $\textit{rm}(c)$. In this case, $a$ is the only value of priority smaller than $p_2$ and it satisfies this property. Applying a similar reasoning for all the remaining values, it can be proved that this execution is correct.

\begin{algorithm}[t]
\footnotesize{
\KwIn {A data-differentiated sequential execution $e$}
\KwOut{$\mathsf{true}$ iff $e\in \seqPQ$}

@@

\Else
{
    \If {$\exists\ x \in \textit{values}(e)$ such that $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ holds}
    {
        \KwRet $\textit{Check-PQ-Seq}(e \setminus x)$\;
    }
    \Else {\KwRet $\mathsf{false}$\;}
}}
\caption{$\textit{Check-PQ-Seq}$}
\label{alg:seq_check}
\end{algorithm}

Formally, the selected operations depend on the following set of predicates on executions:
%{\small
\begin{align*}
& \mathsf{Has\text{-}EmptyRemoves}(e)=\mathsf{true} \mbox{ iff  $e$ contains a $\textit{rm}(\textit{empty})$-operation} \hspace{1cm}\\
& \mathsf{Has\text{-}UnmatchedMaxPriority}(e)=\mathsf{true} \mbox{ iff $p\in \textit{unmatched-priorities}(e)$ for a maximal $p$} 
%&\hspace{4mm}\mbox{$p\in priorities(e)$}
\end{align*}%}
where $\textit{priorities}(e)$, resp., $\textit{unmatched-priorities}(e)$, is the set of priorities occurring in $\textit{put}$ operations of $e$, resp., in $\textit{put}$ operations of $e$ for which there is no $\textit{rm}$ operation removing the same value. We call the latter \emph{unmatched} put operations. A put operation which is not unmatched is called \emph{matched}. For simplicity, we consider the following syntactic sugar $\mathsf{Has\text{-}MatchedMaxPriority}(e)=\neg \mathsf{Has\text{-}EmptyRemoves}(e)\land \neg \mathsf{Has\text{-}UnmatchedMaxPriority}(e)$. By an abuse of notation, we assume  $\mathsf{Has\text{-}UnmatchedMaxPriority}(e) \Rightarrow \neg \mathsf{Has\text{-}EmptyRemoves}(e)$ (this is sound by the order of the conditionals in $\textit{Check-PQ-Seq}$).

The predicates defining the correctness of the selected operations are defined as follows:
%{\small
\begin{align*}
&\mathsf{EmptyRemove\text{-}Seq}(e,o)=\mathsf{true} \mbox{ iff  $e= u\cdot o\cdot v$ and $\textit{matched}(u)$} \\
&\mathsf{UnmatchedMaxPriority\text{-}Seq}(e,x)=\mathsf{true}  \mbox{ iff  $e= u\cdot \textit{put}(x,p)\cdot v$, $p\not\prec \textit{priorities}(u\cdot v)$,} \\
&\hspace{6.4cm}p\not\in \textit{priorities}(v) \\
&\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)=\mathsf{true} \mbox{ iff  $e= u\cdot \textit{put}(x,p)\cdot v\cdot \textit{rm}(x)\cdot w$, $\textit{matched}_\prec(u\cdot v,p)$,} \\
&\hspace{4.3cm}\mbox{$p\not\preceq \textit{unmatched-priorities}(u\cdot v\cdot w)$, $p\not\prec \textit{priorities}(u\cdot v\cdot w)$,} \\
&\hspace{4.3cm}\mbox{and  $p\not\in \textit{priorities}(v\cdot w)$}
\end{align*}
%}
where $p\prec \textit{priorities}(e)$ when $p\prec p'$ for some $p'\in \textit{priorities}(e)$ (and similarly for $p\prec \textit{unmatched-priorities}(e)$ or $p\preceq \textit{unmatched-priorities}(e)$),
$\textit{matched}_\prec(e,p)$ holds when each value with priority strictly smaller than $p$ is removed in $e$, and $\textit{matched}(e)$ holds when $\textit{matched}_\prec(e,p)$ holds for each $p\in\mathbb{P}$. Compared to the example presented at the beginning of the section, these predicates take into consideration that multiple values with the same priority are removed in FIFO order: the predicate $\mathsf{MatchedMaxPrioritySeq}(e,x)$ holds when $x$ is the last value with priority $p$ added in $e$.

When $o$ is an $\textit{rm}(\textit{empty})$-operation, $e\setminus o$ is the maximal subsequence of $e$ which doesn't contain $o$. For an execution $e$, $\textit{values}(e)$ is the set of values  in call/return actions of $e$.

@@

The extension of $\textit{Check-PQ-Seq}$ to concurrent executions, checking whether they are linearizable w.r.t. $\seqPQ$, is obtained by replacing every predicate $\Gamma\mathsf{\text{-}Seq}$ with
\begin{align*}
\Gamma\mathsf{\text{-}Conc}(e,\alpha) = \mathsf{true}\mbox{ iff there is a sequential execution $s$ such that $e\sqsubseteq s$ and $\Gamma\mathsf{\text{-}Seq}(s,\alpha)$}
\end{align*}
for each $\Gamma\in \{\mathsf{EmptyRemove}, \mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$. The obtained procedure is denoted by 
$\textit{Check-PQ-Conc}$ (recursive calls are modified accordingly).

@@

The following lemma states the correctness of $\textit{Check-PQ-Conc}$. Completeness follows easily from the properties of $\seqPQ$. If $\textit{Check-PQ-Conc}(e) = \textit{false}$, then there exists a set $D$ of values s.t. either  $\mathsf{EmptyRemove\text{-}Conc}(e \vert D)$ is false, or $\mathsf{UnmatchedMaxPriority\text{-}Conc}(e \vert D,x)$ is false for all the values $x$ of maximal priority that are not removed (and there exists at least one such value), or $\mathsf{MatchedMaxPriority\text{-}Conc}(e \vert D,x)$ is false for all the values $x$ of maximal priority (and these values are all removed in $e \vert D$). It can be easily seen that we get $e \vert D\not\sqsubseteq \seqPQ$ in all cases, which by the closure under projection of $\seqPQ$ implies, $e \not\sqsubseteq \seqPQ$ (since every linearization of $e$ includes as a subsequence a linearization of $e \vert D$).

\begin{restatable}{lemma}{ConCheckEPQIsCorrect}
\label{lemma:con-check-EPQ is correct}
$\textit{Check-PQ-Conc}(e)=\mathsf{true}$ iff $e \sqsubseteq \seqPQ$, for every data-differentiated $e$.
\end{restatable}

@@

%This completes the proof of this lemma.
%
%
%\qed
%\end {proof}
\begin{figure}[t]
  \centering
  \includegraphics[width=1 \textwidth]{PIC-HIS-EPQ1-TwoHis-2.pdf}
%\vspace{-10pt}
  \caption{The process of obtaining linearization of $e$}
  \label{fig:concurrent execution for EPQ1}
\end{figure}

### 2017-07-10 final/introduction.tex
half of the paper

Modern computer software is increasingly concurrent. Interactive applications
and services necessitate reactive asynchronous operations to handle requests
immediately as they happen, rather than waiting for long-running operations to
complete. Furthermore, as processor manufacturers approach clock-speed limits,
performance improvements are more-often achieved by parallelizing operations
across multiple processor cores.

@@

 Multithreaded software is typically built with specialized “concurrent
  objects” like atomic integers, queues, maps, priority queues. These objects’ methods are
  designed to confom to better established sequential specifications, a property known as \emph{linearizability}~\cite{journals/toplas/HerlihyW90},
  despite being optimized to avoid blocking and exploit parallelism, e.g.,~by
  using atomic machine instructions like compare-and-swap. Intuitively, linearizability asks that every individual operation appears to take place instantaneously at some point between its invocation and its return. Verifying linearizability is intrinsically hard, and undecidable in general~\cite{conf/esop/BouajjaniEEH13}. % (even for finite-state implementations).
However, recent work~\cite{DBLP:conf/icalp/BouajjaniEEH15} has shown that for particular classes of objects,
i.e., registers, mutexes, queues, and stacks, the problem of verifying linearizability becomes decidable (for finite-state implementations).

In this paper, we consider another important object, namely the priority queue, which is essential for applications such as task scheduling and discrete event simulation. Numerous implementations have been proposed in the research literature, e.g.,~\cite{DBLP:conf/ppopp/AlistarhKLS15,DBLP:conf/wdag/CalciuMH14,DBLP:conf/opodis/LindenJ13,DBLP:conf/podc/ShavitZ99,DBLP:conf/ipps/ShavitL00}, and concrete implementations exist in many modern languages like C++ or Java.
%
%This object is much less studied compared to the other ones in the verification literature.
%
Priority queues are collections providing $\textit{put}$ and $\textit{rm}$ methods for adding and removing values. Every value added is associated to a priority and a remove operation returns a minimal priority value. For generality, we consider a partially-ordered set of priorities. Values with incomparable priorities can be removed in any order, and values having the same priority are removed in the FIFO order. Implementations like the PriorityBlockingQueue in Java where same priority values are removed in an arbitrary order can be modeled in our framework by renaming equal priorities to incomparable priorities (while preserving the order constraints).

Compared to previously studied collections like stacks and queues, the main challenge in dealing with priority queues is that the order in which values are removed is not fixed by the happens-before between add/remove operations (e.g., in the case of a queue, the values are removed in the order in which they were inserted), but by parameters of the $\textit{put}$ operations (the priorities) which come from an unbounded domain. For instance, the sequential behavior $\textit{put}(a,p_1)\cdot \textit{put}(b,p_3)\cdot \textit{put}(c,p_2)\cdot \textit{rm}(a,p_1)\cdot \textit{rm}(c,p_2)$ where the priority $p_1$ is less than $p_2$ which is less than $p_3$, is not admitted neither by the regular queue nor the stack.

@@

Following the approach in~\cite{DBLP:conf/icalp/BouajjaniEEH15}, we give a characterization of concurrent priority queue behaviors violating linearizability in terms of automata. This characterization enables a reduction of checking linearizability for arbitrary implementations to reachability or invariant checking, and implies decidability for checking linearizability of finite-state implementations. However, differently from the case of stacks and queues where finite-state automata are sufficient, we show that the case of priority queues needs \emph{register automata} where registers are used to store and compare priorities.

This characterization is obtained in several steps. We first define a recursive procedure which recognizes valid sequential executions which is then extended to recognize linearizable concurrent executions. Intuitively, for an input execution $e$, this procedure handles values occurring in $e$ one by one, starting with values of maximal priority (which are to be removed the latest). For each value $x$, it checks whether $e$ satisfies some property ``local'' to that value, i.e., which is agnostic to how the operations adding or removing other values are ordered between them (w.r.t. the happens-before), other than how they are ordered w.r.t. the operations on $x$. When this property holds, the same is done for the rest of the execution, without the operations on $x$. This procedure works only for executions where a value is added at most once, but this is not a limitation for \emph{data-independent} implementations whose behavior doesn't depend on the values that are added or removed. In fact, all the implementations that we are aware of are data-independent.

@@

The remainder of this article is organized as follows.
Section~\ref{sec:priority queue and data-independence} describes the priority queue ADT, lists several semantic properties like data-independence that are satisfied by implementations of this ADT, and formalizes the notion of linearizability.
Sections~\ref{sec:checking inclusion by recursive procedure} defines a recursive procedure for checking linearizability of concurrent priority queue behaviors.
Section~\ref{sec:co-regular of extended priority queues} gives an automata characterization of the violations to linearizability, and
Section~\ref{sec:related} discusses related work. {\color {red}Some proofs and detailed construction of register automata are deferred to the long version \cite{CONCUR2017Ahmed}.} 
%, and
%Section~\ref{sec:conclusion} gives concluding remarks.

>>>>

%Modern computer software is increasingly concurrent. Interactive applications
%and services necessitate reactive asynchronous operations to handle requests
%immediately as they happen, rather than waiting for long-running operations to
%complete. Furthermore, as processor manufacturers approach clock-speed limits,
%performance improvements are more-often achieved by parallelizing operations
%across multiple processor cores.

@@

 Multithreaded software is typically built with specialized “concurrent
  objects” like atomic integers, queues, maps, priority queues. These objects’ methods are
  designed to confom to better established sequential specifications, a property known as \emph{linearizability}~\cite{journals/toplas/HerlihyW90},
  despite being optimized to avoid blocking and exploit parallelism, e.g.,~by
  using machine instructions like compare-and-swap. Intuitively, linearizability asks that every individual operation appears to take place instantaneously at some point between its invocation and its return. Verifying linearizability is intrinsically hard, and undecidable in general~\cite{conf/esop/BouajjaniEEH13}. % (even for finite-state implementations).
However, recent work~\cite{DBLP:conf/icalp/BouajjaniEEH15} has shown that for particular objects,
e.g., registers, mutexes, queues, and stacks, the problem of verifying linearizability becomes decidable (for finite-state implementations).

In this paper, we consider another important object, namely the priority queue, which is essential for applications such as task scheduling and discrete event simulation. Numerous implementations have been proposed in the research literature, e.g.,~\cite{DBLP:conf/ppopp/AlistarhKLS15,DBLP:conf/wdag/CalciuMH14,DBLP:conf/opodis/LindenJ13,DBLP:conf/podc/ShavitZ99,DBLP:conf/ipps/ShavitL00}, and concrete implementations exist in many modern languages like C++ or Java.
%
%This object is much less studied compared to the other ones in the verification literature.
%
Priority queues are collections providing $\textit{put}$ and $\textit{rm}$ methods for adding and removing values. Every added value is associated to a priority and a remove operation returns a minimal priority value. For generality, we consider a partially-ordered set of priorities. Values with incomparable priorities can be removed in any order, and values having the same priority are removed in the FIFO order. Implementations like the PriorityBlockingQueue in Java where same priority values are removed in an arbitrary order can be modeled in our framework by renaming equal priorities to incomparable priorities (while preserving the order constraints).

Compared to previously studied collections like stacks and queues, the main challenge in dealing with priority queues is that the order in which values are removed is not fixed by the happens-before between add/remove operations (e.g., in the case of queues, values are removed in the order in which they were inserted), but by parameters of the $\textit{put}$ operations (the priorities) which come from an unbounded domain. For instance, the sequential behavior $\textit{put}(a,p_1)\cdot \textit{put}(b,p_3)\cdot \textit{put}(c,p_2)\cdot \textit{rm}(a,p_1)\cdot \textit{rm}(c,p_2)$ where the priority $p_1$ is less than $p_2$ which is less than $p_3$, is not admitted neither by the regular queue nor the stack.

@@

We give a characterization of concurrent priority queue behaviors violating linearizability in terms of automata. This characterization enables a reduction of checking linearizability for arbitrary implementations to reachability or invariant checking, and implies decidability for checking linearizability of finite-state implementations. While linearizability violations for stacks and queues can be described using finite-state automata~\cite{DBLP:conf/icalp/BouajjaniEEH15}, the case of priority queues requires \emph{register automata} where registers are used to store and compare priorities.

This characterization is obtained in several steps. We define a recursive procedure that recognizes valid sequential executions, which is then extended to recognize linearizable concurrent executions. Intuitively, for an execution $e$, this procedure deals with values occurring in $e$ one by one, starting with values of maximal priority (to be removed the latest). For each value $x$, it checks whether $e$ satisfies some property ``local'' to that value, i.e., which is agnostic to how the operations adding or removing other values are ordered between them (w.r.t. the happens-before), other than how they are ordered w.r.t. the operations on $x$. When this property holds, the procedure is applied recursively on the rest of the execution, without the operations on $x$. This procedure works only for executions where a value is added at most once, but this is not a limitation for \emph{data-independent} implementations whose behavior doesn't depend on the values that are added or removed. In fact, all the implementations that we are aware of are data-independent.

@@

The remainder of this article is organized as follows.
Section~\ref{sec:priority queue and data-independence} describes the priority queue ADT, lists several semantic properties like data-independence, and recalls the notion of linearizability.
Section~\ref{sec:checking inclusion by recursive procedure} defines a recursive procedure for checking linearizability of concurrent priority queue behaviors.
Section~\ref{sec:co-regular of extended priority queues} gives an automata characterization of the violations to linearizability, and
Section~\ref{sec:related} discusses related work. Detailed proofs and constructions can be found in the extended version~\cite{CONCUR2017Ahmed}. 
%, and
%Section~\ref{sec:conclusion} gives concluding remarks.
